The Development of a Collaborative Tool to Teach Debugging

### 2020-12-04 paper/paper.tex
Other example programs and some more writing

@@

Similarly to research on the value of teaching debugging, research
into how to best teach debugging is self-admittedly sparse.  Chan et
al. allow that ``in general research on how to improve debugging is
sporadic''---an observation that leads them to research a framework to
reduce the complexity of teaching
debugging\cite{10.1145/3286960.3286970}.  To organize their framework,
they split debugging knowledge into 5 categories: \textit{Domain},
\textit{System}, \textit{Procedural}, \textit{Strategic}, and
\textit{Experiential}.  They then review different debugging tools and
teaching aids---from those that involve writing code to games---and
map tools to the knowledge areas they seek to address.  After an
evaluation of a host of different tools, they claim to find a few
significant faults in current debugging teaching platforms.  The
collaborative debugger primarily seeks to address the lack of
back-tracing ability/coverage found in their research.

@@

Research into the value of collaborative programming is overwhelmingly
positive.  McDowell et. al. found that not only does collaborative
programming significantly boost student confidence and the retention
of students in computer science majors, but that it demonstrably
improves student's work \cite{10.1145/1145287.1145293}.  These
benefits of confidence and correctness are present when paired and
non-paired students are given identical assignments
\cite{10.1145/1026487.1008043}, further indicating that the simple act
of collaboration definitively benefits computer science students.\par

It seems that the benefits in confidence that result from teaching
debugging should be magnified by teaching debugging collaboratively.
There are similarities between the introductory nature of teaching
fundemental debugging skills and the nature of teaching fundemental
programming skills covered in the classes studied in
\cite{10.1145/1026487.1008043} and \cite{10.1145/1145287.1145293}.
Introductory programming classes typically use a specific programming
language in order to introduce widely applicable programming
principles.  By using a specific collaborative platform to introduce
debugging principles, students may realize the same benefits in both
confidence \textit{and} correctness that they do from pair programming
in introductory comptuer science classes.\par

The collaborative debugger aims to provide such a platform.

@@

TODO: write about glitch, repl.it, etc.

Debuggers exist at an intersection of tools and skills similar to
programming languages themselves. By becoming familiar with a specific
debugger, students learn techniques and paradigms necessary to use all
debuggers effectively.

@@

Docker containers are created using a Dockerfile.  The Dockerfile used
to create the container image for the RR Message
Server is shown below:\\

@@

The Pod configuration schema for most Pods in the collaborative
debugger is fairly generic.  It consists of a \lstinline{name}, an
\lstinline{image} sourced from the container registry, and in the case
of pods that need to interact with a load balancer, an
\lstinline{app}.\\

\begin{lstlisting}[language=YAML,basicstyle=\linespread{0.5}\ttfamily\small,caption={RR Debug Session Schema},captionpos=b]
apiVersion: v1
kind: Pod
metadata:
  name: rr-translation
  labels:
    purpose: translate-rr
spec:
  containers:
  - name: rr-test-container
    image: example-container-registry.com/sproj/...
    securityContext:
      capabilities:
        add:
        - SYS_PTRACE
  restartPolicy: OnFailure
\end{lstlisting}

@@

Services are the first components of the collaborative debugger
created after the cluster is initiallized.  The three Services used by
the debugger are all statically created.  Like statically created
Pods, schemas that define Services are applied manually using the
\lstinline{kubectl apply} command.\\

\begin{lstlisting}[language=YAML,basicstyle=\linespread{0.5}\ttfamily\small,caption={Frontend Load Balancer Schema},captionpos=b]
apiVersion: v1
kind: Service
metadata:
  name: frontend-load-balancer
spec:
  selector:
    app: frontend
  type: LoadBalancer
  ports:
    - port: 3000
    targetPort: 3000
\end{lstlisting}

@@

The RR Message server is remarkably simple, consisting of just 3
important functions:\\

\begin{lstlisting}[language=Python,basicstyle=\linespread{0.5}\ttfamily\small,caption={RR Message Server},captionpos=b]
@sio.on('join_channel')
def join_channel(sid, data):
    sio.enter_room(sid, data['channel'])

@@

The process for building RR Debug Session example images is differs
slightly from other images used by the collaborative debugger.
Currently, three example RR Debug Session images are availiable for
use.  Example container images are based on the primary RR Debug
Session image, RR Translation.  The build process for this image,
based on the lasted official Ubuntu container image, installs rr,
python, and all necessary packages as well as the app that will run on
the final image, \lstinline{rrtranslation.py}.\\

\begin{lstlisting}[language=Dockerfile,basicstyle=\small,caption={RR Debug Session Hash Example---Dockerfile},captionpos=b]
FROM example-container-registry.com/sproj/...

@@

The Dockerfile shown above is for a simple hash program RR Debug
Session example.  The build process copies over any necessary files,
as well as a startup script:\\

\begin{lstlisting}[basicstyle=\linespread{0.5}\ttfamily\small,caption={Example Startup Script},captionpos=b]
gcc -g -o hash hash.c
rr record ./hash
python3 rrtranslation.py $1
\end{lstlisting}

@@

The RR Translation program's socket.io client instance begins by
connecting to the RR Message Server.  It immediately emits a
\lstinline{'join_channel'} message, using the channel passed in from
the Kubernetes API call on it's creation. \\

\begin{lstlisting}[language=Python,basicstyle=\linespread{0.5}\ttfamily\small,caption={RR Translation Main},captionpos=b]
if __name__ == '__main__':
    channel = sys.argv[1]
    sio.connect('http://rr-message-server-load-balancer:8000')
    sio.emit('join_channel', {'channel': channel})
\end{lstlisting}

After joining the appropriate channel, \lstinline{sio} waits to
recieve an \lstinline{'rr_command'}.  The first part of
\lstinline{sio}'s \lstinline{'rr_command'} event handler is shown below:\\

\begin{lstlisting}[language=Python,basicstyle=\linespread{0.5}\ttfamily\small,caption={RR Command Event Handler},captionpos=b]
@sio.on('rr_command')
def on_rr_command(data):

@@

When the program starts, an instance of the \lstinline{RRInterface}
class, \lstinline{rri}, is initialized in the same scope as
\lstinline{sio}.  \lstinline{RRInterface} contains an instance of the
\lstinline{GDBController} class from pygdbmi to interface with rr and
control the rr subprocess, as well as a collection of methods to parse
rr output, the three most important of which are shown below: \\

\begin{lstlisting}[language=Python,basicstyle=\linespread{0.5}\ttfamily\small,caption={RRInterface},captionpos=b]
def get_full_rr_response(self, command):
    response = self.gdbmi.write(command)
    while(not(self.end(response) or self.exited(response))):
        response.extend(self.get_rr_response())
    return response

@@

Effort is taken in the design of the RR Debug Session and RR Message
Server to ensure students recieve a pleasant and near-identical
experience to debugging locally in the future.  The aim is to provide
students with an experience they can directly reference when using rr
or gdb in the future, as well as one which is seamless to encourage
the acquisition of debugging skills.  The system of inheritance used
to create example debug session pods is designed to be as simple as
possible, though a more fully featured frontend could simplify this
process further in the future.

@@

The Frontend/API Server uses Yarn \cite{yarn} to manage packages.
When a Frontend/API Server Pod is deployed, the Pod's startup script
uses the \lstinline{yarn start} and \lstinline{yarn start-api}
commands to start the frontend server and API server respectively.
These scripts are defined in the server's \lstinline{package.json}
configuration file:\\

\begin{lstlisting}[basicstyle=\linespread{0.5}\ttfamily\small,caption={Frontend/API Server Configuration},captionpos=b]
  "start": "react-scripts start",
  "start-api": "cd api && flask run --no-debugger",
  .
  .
  .
  "proxy": "http://localhost:5000",
\end{lstlisting}

@@

Below is the route handler for the \lstinline{'/new'} URL, as well as
the instantiation of \lstinline{TranslationPodManager} and
\lstinline{UserManager}.\\

\begin{lstlisting}[language=Python,basicstyle=\linespread{0.5}\ttfamily\small,caption={API Server New RR Debug Session Event Handler},captionpos=b]
tpm = podmanager.TranslationPodManager(
                 url='mongodb://database-load-balancer',
                 port=27017)
um = usermanager.UserManager(
                 url='mongodb://database-load-balancer',
                 port=27017)
app = Flask(__name__)

@@

% \begin{lstlisting}[basicstyle=\linespread{0.5}\ttfamily\small,caption={Frontend/API Server Configuration},captionpos=b]
% users:            pods:               examples:
%   _id ObjectId      _id ObjectId        _id ObjectId
%   name String       channel String      name String
%   pods Array        users Array         container String

@@

To dynamically create a Pod, \lstinline{create_pod} gets the container
registry URL of the image to base the Pod on, creates the new Pod,
binds users to the Pod, and returns the channel the Pod was created
with.  The first part of the function is shown below:\\

\begin{lstlisting}[language=Python,basicstyle=\linespread{0.5}\ttfamily\small,caption={Pod Creation 1},captionpos=b]
def create_pod(self, names, program):
  examples = self.get_examples()
  image = None
  try:
    image = examples[program]
  except:
    raise NoSuchExampleError
\end{lstlisting}

\lstinline{create_pod} calls \lstinline{TranslationPodManager}'s
\lstinline{get_examples} method to get a list of RR Debug Session
example image names and container registry URLs from the database.  In
the event that the user has somehow pased a spurious image name or has
passed a name when there are no images, an exception is raised.
Otherwise, a RR Debug Session Pod schema is created using the image:\\

\begin{lstlisting}[language=Python,basicstyle=\linespread{0.5}\ttfamily\small,caption={Pod Creation 2},captionpos=b]
  dep = {'apiVersion': 'v1',
         'kind': 'Pod',
         'metadata': {'labels':
                      {'purpose': 'translate-rr'}},
         'spec': {'containers':
                  [{'image': image,
                    'name': 'rr-test-container',
                    'command': ['sh'],
                    'args': ['startup.sh'],
                    'securityContext':
                    {'capabilities':
                     {'add': ['SYS_PTRACE']}}}],
                    'restartPolicy': 'OnFailure'}}
\end{lstlisting}

This schema, represented as a dictionary in Python, is identical to
the YAML representation of the RR Debug Session Pod creation schema
shown in (\ref{podschema}) with the exception of \lstinline{args}
which will be updated further in a later step.  Finally,
\lstinline{create_pod} does the
work necessary to dynamically create a Pod:\\

\begin{lstlisting}[language=Python,basicstyle=\linespread{0.5}\ttfamily\small,caption={Pod Creation 3},captionpos=b]
  channel = self.dbc.add_pod(
            self.dbc.get_userids_by_name(names))
  dep['metadata']['name'] = channel
  dep['spec']['containers'][0]['args'].append(channel)
  resp = self.v1.create_namespaced_pod(
         body=dep, namespace='default')
  # Wait to return the channel until the pod is live
  # and read to recieve incomming communication
  status = self.v1.read_namespaced_pod_status(
           channel, 'default').status.container_statuses
  while status == None or status[0].state.running == None:
    time.sleep(1)
    status = self.v1.read_namespaced_pod_status(
             channel, 'default').status.container_statuses
  return channel
\end{lstlisting}

>>>>

\lstset{frame=tb,
  aboveskip=8mm,
  columns=flexible,
  numbers=none,
  tabsize=3
}

@@

Similarly to research on the value of teaching debugging, research
into how to best teach debugging is sparse.  Chan et al. allow that
``in general research on how to improve debugging is sporadic''---an
observation that leads them to research a framework to reduce the
complexity of teaching debugging\cite{10.1145/3286960.3286970}.  To
organize their framework, they split debugging knowledge into 5
categories: \textit{Domain}, \textit{System}, \textit{Procedural},
\textit{Strategic}, and \textit{Experiential}.  They then review
different debugging tools and teaching aids---from those that involve
writing code to games---and map tools to the knowledge areas they seek
to address.  After an evaluation of a host of different tools, they
claim to find a few significant faults in current debugging teaching
platforms.  The collaborative debugger primarily seeks to address the
lack of back-tracing ability/coverage found in their research.

@@

Research into the value of collaborative programming is overwhelmingly
positive.  McDowell et. al. found that not only does collaborative
programming significantly boost student confidence and the retention
of students in computer science majors, but that it demonstrably
improves student's work \cite{10.1145/1145287.1145293}.  These
benefits of confidence and correctness are present when paired and
non-paired students are given identical assignments
\cite{10.1145/1026487.1008043}, further indicating that the simple act
of collaboration definitively benefits computer science students.
\par

It seems that the benefits in confidence that result from teaching
debugging should be magnified by teaching debugging collaboratively.
There are similarities between the introductory nature of teaching
fundemental debugging skills and the nature of teaching fundemental
programming skills covered in the classes studied in
\cite{10.1145/1026487.1008043} and \cite{10.1145/1145287.1145293}.
Introductory programming classes typically use a specific programming
language in order to introduce widely applicable programming
principles.  By using a specific collaborative platform to introduce
debugging principles, students may realize the same benefits in both
confidence \textit{and} correctness that they do from pair programming
in introductory comptuer science classes.
\par

Debuggers exist at an intersection of tools and skills similar to
programming languages themselves. By becoming familiar with a specific
debugger, students may learn techniques and paradigms necessary to use
all debuggers effectively.  The collaborative debugger aims to provide
the optimal platform for students to learn debugging skills.

\section{Requirments}

In order to provide the best platform for both teachers and students,
the collaborative debugger must fufill two key requirements:

\begin{enumerate}
\item It must encourage collaboration through a seamless experience.
\item It must make it easier to enhance students' understanding of
  debugging and programs being debugged.
\end{enumerate}

\subsection{Encouraging Collaboration}

A large number of tools exist to facilitate collaborative programming.
COVID19 has greatly increased the demand for tools that not only make
collaboration easier, but make it easy remotely.  The tool that was
most influential in the design of the collaborative debugger,
Replit\cite{replit}, enables remote collaborative programming.

Replit provides a simple browser-based IDE for over 50 languages.
Programmers choose a language and can edit and run code
collaboratively inside the Replit webapp.  Similarly to other
collaborative text-editors, input and output from all users is synced.
\par

The collaborative debugger aims to provide a similar experience to
Replit.  Students should be able to start a debugging session, easily
invite other students, and interact with the debugger in a way that
makes it easy to share their knowledge.  By allowing multiple users to
interact with the same debugging session together in real-time, the
collaborative debugger lets students share experience and work through
problems together.  The debugger also makes it easy for teachers to
demonstrate debugging techniques remotely to multiple students, any of
whom can also interact with the debugger.  Since debugging sessions
are hosted remotely, users can start a session in one location and
resume it later from a different computer.
\par

\subsection{Enhancing Student's Understanding}

Apart from the benefits to understanding that are realized through
collaboration, the collaborative debugger should enhance students'
understanding of debugging in ways that a traditional debugger cannot.
It aims to do this by allowing teachers to design and distribute
lessons in the form of debugging sessions.  Each debugging session is
a lesson consisting of a deterministic recording of program execution,
which students can debug repeatedly in order to build debugging skills
and to learn about the execution process.
\par

The aim is to provide students with an experience they can directly
reference when using rr or gdb in the future, as well as one which is
seamless to encourage the acquisition of debugging skills.

@@

@@

Docker containers are created using a Dockerfile.  The Dockerfile used
to create the container image for the RR Message
Server is shown below:

@@

The Pod configuration schema for most Pods in the collaborative
debugger is fairly generic.  It consists of a \lstinline{name}, an
\lstinline{image} sourced from the container registry, and in the case
of pods that need to interact with a load balancer, an
\lstinline{app}.

\begin{lstlisting}[language=YAML,basicstyle=\linespread{0.5}\ttfamily,caption={RR Debug Session Schema},captionpos=b]
apiVersion: v1
kind: Pod
metadata:
  name: rr-translation
  labels:
    purpose: translate-rr
spec:
  containers:
  - name: rr-test-container
    image: example-container-registry.com/sproj/...
    securityContext:
      capabilities:
        add:
        - SYS_PTRACE
  restartPolicy: OnFailure
\end{lstlisting}

@@

Services are the first components of the collaborative debugger
created after the cluster is initiallized.  The three Services used by
the debugger are all statically created.  Like statically created
Pods, schemas that define Services are applied manually using the
\lstinline{kubectl apply} command.

\begin{lstlisting}[language=YAML,basicstyle=\linespread{0.5}\ttfamily,caption={Frontend Load Balancer Schema},captionpos=b]
apiVersion: v1
kind: Service
metadata:
  name: frontend-load-balancer
spec:
  selector:
    app: frontend
  type: LoadBalancer
  ports:
    - port: 3000
    targetPort: 3000
\end{lstlisting}

@@

The RR Message server is remarkably simple, consisting of just 3
important functions:

\begin{lstlisting}[language=Python,basicstyle=\linespread{0.5}\ttfamily,caption={RR Message Server},captionpos=b]
@sio.on('join_channel')
def join_channel(sid, data):
    sio.enter_room(sid, data['channel'])

@@

The process for building RR Debug Session example images is differs
slightly from other images used by the collaborative debugger.
Currently, three example RR Debug Session images are availiable for
use.  Example container images are based on the primary RR Debug
Session image, RR Translation.  The build process for this image,
based on the lasted official Ubuntu container image, installs rr,
python, and all necessary packages as well as the app that will run on
the final image, \lstinline{rrtranslation.py}.

\begin{lstlisting}[language=Dockerfile,basicstyle=,caption={RR Debug Session Hash Example---Dockerfile},captionpos=b]
FROM example-container-registry.com/sproj/...

@@

The Dockerfile shown above is for a simple hash program RR Debug
Session example.  The build process copies over any necessary files,
as well as a startup script:

\begin{lstlisting}[basicstyle=\linespread{0.5}\ttfamily,caption={Example Startup Script},captionpos=b]
gcc -g -o hash hash.c
rr record ./hash
python3 rrtranslation.py $1
\end{lstlisting}

@@

The RR Translation program's socket.io client instance begins by
connecting to the RR Message Server.  It immediately emits a
\lstinline{'join_channel'} message, using the channel passed in from
the Kubernetes API call on it's creation.

\begin{lstlisting}[language=Python,basicstyle=\linespread{0.5}\ttfamily,caption={RR Translation Main},captionpos=b]
if __name__ == '__main__':
    channel = sys.argv[1]
    sio.connect('http://rr-message-server-load-balancer:8000')
    sio.emit('join_channel', {'channel': channel})
\end{lstlisting}

After joining the appropriate channel, \lstinline{sio} waits to
recieve an \lstinline{'rr_command'}.  The first part of
\lstinline{sio}'s \lstinline{'rr_command'} event handler is shown below:

\begin{lstlisting}[language=Python,basicstyle=\linespread{0.5}\ttfamily,caption={RR Command Event Handler},captionpos=b]
@sio.on('rr_command')
def on_rr_command(data):

@@

When the program starts, an instance of the \lstinline{RRInterface}
class, \lstinline{rri}, is initialized in the same scope as
\lstinline{sio}.  \lstinline{RRInterface} contains an instance of the
\lstinline{GDBController} class from pygdbmi to interface with rr and
control the rr subprocess, as well as a collection of methods to parse
rr output, the three most important of which are shown below:

\begin{lstlisting}[language=Python,basicstyle=\linespread{0.5}\ttfamily,caption={RRInterface},captionpos=b]
def get_full_rr_response(self, command):
    response = self.gdbmi.write(command)
    while(not(self.end(response) or self.exited(response))):
        response.extend(self.get_rr_response())
    return response

@@

Effort is taken in the design of the RR Debug Session and RR Message
Server to ensure students recieve a pleasant and near-identical
experience to debugging locally in the future.  The system of
inheritance used to create example debug session pods is designed to
be as simple as possible, though a more fully featured frontend could
simplify this process further in the future.

@@

The Frontend/API Server uses Yarn \cite{yarn} to manage packages.
When a Frontend/API Server Pod is deployed, the Pod's startup script
uses the \lstinline{yarn start} and \lstinline{yarn start-api}
commands to start the frontend server and API server respectively.
These scripts are defined in the server's \lstinline{package.json}
configuration file:

\begin{lstlisting}[basicstyle=\linespread{0.5}\ttfamily,caption={Frontend/API Server Configuration},captionpos=b]
  "start": "react-scripts start",
  "start-api": "cd api && flask run --no-debugger",
  .
  .
  .
  "proxy": "http://localhost:5000",
\end{lstlisting}

@@

Below is the route handler for the \lstinline{'/new'} URL, as well as
the instantiation of \lstinline{TranslationPodManager} and
\lstinline{UserManager}.

\begin{lstlisting}[language=Python,basicstyle=\linespread{0.5}\ttfamily,caption={API Server New RR Debug Session Event Handler},captionpos=b]
tpm = podmanager.TranslationPodManager(
                   url='mongodb://database-load-balancer',
                   port=27017)
um = usermanager.UserManager(
                   url='mongodb://database-load-balancer',
                   port=27017)
app = Flask(__name__)

@@

% \begin{lstlisting}[basicstyle=\linespread{0.5}\ttfamily,caption={Frontend/API Server Configuration},captionpos=b]
% users:            pods:               examples:
%   _id ObjectId      _id ObjectId        _id ObjectId
%   name String       channel String      name String
%   pods Array        users Array         container String

@@

To dynamically create a Pod, \lstinline{create_pod} gets the container
registry URL of the image to base the Pod on, creates the new Pod,
binds users to the Pod, and returns the channel the Pod was created
with.  The first part of the function is shown below:

\begin{lstlisting}[language=Python,basicstyle=\linespread{0.5}\ttfamily,caption={Pod Creation 1},captionpos=b]
def create_pod(self, names, program):
  examples = self.get_examples()
  image = None
  try:
    image = examples[program]
  except:
    raise NoSuchExampleError
\end{lstlisting}

\lstinline{create_pod} calls \lstinline{TranslationPodManager}'s
\lstinline{get_examples} method to get a list of RR Debug Session
example image names and container registry URLs from the database.  In
the event that the user has somehow pased a spurious image name or has
passed a name when there are no images, an exception is raised.
Otherwise, a RR Debug Session Pod schema is created using the image:

\begin{lstlisting}[language=Python,basicstyle=\linespread{0.5}\ttfamily,caption={Pod Creation 2},captionpos=b]
  dep = {'apiVersion': 'v1',
         'kind': 'Pod',
         'metadata': {'labels':
                      {'purpose': 'translate-rr'}},
         'spec': {'containers':
                  [{'image': image,
                    'name': 'rr-test-container',
                    'command': ['sh'],
                    'args': ['startup.sh'],
                    'securityContext':
                    {'capabilities':
                     {'add': ['SYS_PTRACE']}}}],
                   'restartPolicy': 'OnFailure'}}
\end{lstlisting}

This schema, represented as a dictionary in Python, is identical to
the YAML representation of the RR Debug Session Pod creation schema
shown in (\ref{podschema}) with the exception of \lstinline{args}
which will be updated further in a later step.  Finally,
\lstinline{create_pod} does the
work necessary to dynamically create a Pod:

\begin{lstlisting}[language=Python,basicstyle=\linespread{0.5}\ttfamily,caption={Pod Creation 3},captionpos=b]
  channel = self.dbc.add_pod(
            self.dbc.get_userids_by_name(names))
  dep['metadata']['name'] = channel
  dep['spec']['containers'][0]['args'].append(channel)
  resp = self.v1.create_namespaced_pod(
         body=dep, namespace='default')
  # Wait to return the channel until the pod is live
  # and read to recieve incomming communication
  status = self.v1.read_namespaced_pod_status(
           channel, 'default').status.container_statuses
  while status == None or status[0].state.running == None:
    time.sleep(1)
    status = self.v1.read_namespaced_pod_status(
             channel, 'default').status.container_statuses
  return channel
\end{lstlisting}
